RedWine by Ameen Syed
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(knitr)
library(ggplot2)
library(dplyr)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("/Users/ameensyed/Downloads")
redwine <- read.csv("wineQualityReds.csv")
```

# Dataset Information

This dataset is about the wine quality of the red variant of the Portuguese 
"Vinho Verde" wine. There are 1599 samples of red wine. The attributes of the 
dataset are fixed acidity (tartaric acid g/dm^3), volatile acidity (acetic acid 
g/dm^3), citric acid (g/dm^3), residual sugar (g/dm^3), 
chlorides (sodium chloride g/dm^3), free sulfur dioxide (mg/dm^3), total sulfur 
dioxide (mg/dm^3), density (g/cm^3), pH, sulphates (potassium sulfate g/dm^3),
alcohol (% by volume), and quality (Output variable. Score from 0 to 10). 

The 11 input variables were taken from physicochemical tests while the output 
variable, Quality, was determined by 3 wine experts. 

#Guiding Question: 
> Which chemical properties influence the quality of red wines?

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
dim(redwine)
redwine_no_x <- subset(redwine,select = -X)
#Dropped X column before summarizing since X is just the ID of each sample
summary(redwine_no_x)
```

There are 13 variables and 1599 observations. Since the variable X is just an
ID for each sample, it was dropped when calculating the summary of the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality)) + geom_histogram()
summary(redwine$quality)
```



First off, let's look at the distribution of the quality of the samples. It is
clear that most wines were of a quality of either 5,6, or 7. I think it will be 
important to see what is the difference between wines of quality 7 vs wines of 
quality 5, because there's an adequate amount of samples for the two quality 
levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(alcohol)) + geom_histogram() + 
  scale_x_continuous(breaks = seq(8,14,0.5), limits = c(8,14))
summary(redwine$alcohol)
```

Most of the alcohol distribution was between 9.25 and 9.75.These values are 
relatively far from the max. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(fixed.acidity)) + geom_histogram(binwidth = 0.25)
ggplot(data = redwine, aes(volatile.acidity)) + geom_histogram(binwidth = 0.1) + scale_x_continuous(breaks = seq(0,1.5,0.1),limits = c(0,1.5))
ggplot(data = redwine, aes(citric.acid)) + geom_histogram() + 
  scale_x_continuous(breaks = seq(0,1,0.1)) 

```

From the histograms of the acids, it appears as though in general, there are 
very few samples with extremely high concentrations of acid. Furthermore, fixed
acidity and volatile acidity have a similar distribution where the largest 
frequencies occur in the lower-middle level of fixed acidity or volatile 
acidity. The distribution of citric acid is more visibly different than the 
other two distributions. There are many more samples with low levels
of citric acid (0-0.1 out of 1) than there are for the other two acids. In 
other words, the median value seems to be further from the max value in citric 
acid than the other two acids. I wonder how the box plots will look like in the
next section. For now, I've computed the summary statistics for citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(redwine$citric.acid)

```

Most samples have a citric acid level of less than 0.5. This statistic confirms
our observation than the distribution of citric acid levels is concentrated on 
the lower levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(pH)) + geom_histogram()
summary(redwine$pH)
```

The distribution of pH is similar to volatile acidity and fixed acidity, except
it is more symmetrical. Citric acid's distribution, however, is different 
because it is more concentrated on the lower section while pH is more symmetric,
being most concentrated in the center. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Now let's look at the distribution of Sulfur Dioxide. 
#Sulfur Dioxide comes in 2 forms in wines, free and bound.
#https://en.wikipedia.org/wiki/Sulfur_dioxide
#Since our Data doesn't have bound forms, I'll create the variable
redwine$bound.sulfur.dioxide <- redwine$total.sulfur.dioxide - 
  redwine$free.sulfur.dioxide
ggplot(data = redwine, aes(free.sulfur.dioxide)) + geom_histogram()+
  scale_x_continuous(limits = c(0,100), breaks = seq(0,100,10))
ggplot(data = redwine, aes(bound.sulfur.dioxide)) + geom_histogram()+
  scale_x_continuous(limits = c(0,100), breaks = seq(0,100,10))
ggplot(data = redwine, aes(total.sulfur.dioxide)) + geom_histogram() +scale_x_continuous(limits = c(0,100), breaks = seq(0,100,10))
summary(redwine$total.sulfur.dioxide)

```

The total sulfur dioxide seems to be distributed in the lower section primarily
also. The highest counts of total sulfur dioxide occur from 10 - 30, and they 
keep getting lower in frequency as the level of total sulfur dioxide increases.
In general, there's more bound sulfur dioxide than free sulfur dioxide. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(residual.sugar)) + geom_histogram() 
ggplot(data = redwine, aes(chlorides)) + geom_histogram()

```

Both these histograms have a similar distribution. That's very peculiar, unless
there is a relationship between them. Let's see if they are the same samples or 
different ones, so we know whether to expect a relationship between the 2 
variables or not. It is also worth noting that they are both concentrated on low 
side with some samples having medium, high, and very high levels (of residual 
sugar and chlorides respectively). Let's look more at the high-value samples.

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(redwine,residual.sugar > 8)
subset(redwine, chlorides > 0.3)
```

It appears that they are different samples. So, it is just a coincidence that
the distribution of the histograms look similar. 

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 samples with 12 features (fixed acidity, volatile acidity, citric
acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, 
density, pH, sulphates, alcohol, and quality). 

Some Observations:
- Most wine samples have a quality of 5.
- The median alcohol content was 10.2.
- The distribution of pH matches the distribution of volatile acidity.
- The median total sulfur dioxide level is 38.
- Most wine samples have residual sugar less than 8 and chlorides less than 0.3.


### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is the quality and how other factors relate to it.
I think there can be a predictive model built from some combination of other 
features for determining an optimal wine sample. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other features like acidity, sulfur dioxide, alcohol, and sugar all combine in 
particular amounts to make the ideal wine sample. My goal is to determine which 
features are best for predicting the quality of wine samples.

### Did you create any new variables from existing variables in the dataset?

Yes, I created the Bound Sulfur Dioxide variable, based on my research. I did
this by subtracting Free Sulfur Dioxide from Total Sulfur Dioxide. I think 
Bound Sulfur Dioxide will serve as another helpful factor for determining a 
relationship in the Quality of wine samples.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Yes, I think it was unusual how the distribution of residual sugar and chlorides
was very similar, even though the samples which had the higher values in one 
had no correlation to the higher values in the other. I think it's a very 
strange coincidence that this happend in a collection of 1599 samples. 

I also think it's interesting how the pH level distribution was nearly 
symmetrical. This will make it easier to determine the relationship between
quality and acidity.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
ggplot(data = redwine, aes(quality,alcohol)) + geom_jitter(alpha = 0.2) 
ggplot(data = redwine, aes(group=quality, x = quality, y =  alcohol)) + 
  geom_boxplot() 
summary(redwine$alcohol)
cor(redwine$quality, redwine$alcohol)

```

Alcohol content seems to have a relationship with quality. If we look at the 
scatter plot, there's a heavy concentration of low alcohol content with a 
quality of 5. The alcohol content in quality 6 samples is more scattered out
on top. Quality 7 samples has alcohol content even more scattered out towards
the top. And quality 8 samples has mostly alcohol content greater than 12. These
findings are further reinforced by the fact that the boxes seem to be going 
higher and higher starting from 5 going to 8, with 8 having the highest median.
And since the correlation coefficient of alcohol and quality is 0.476, we can
know for sure that there is somewhat of a relationship with the two variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality,citric.acid)) + geom_jitter(alpha = 0.2)
ggplot(data = redwine, aes(group=quality, x = quality, y =  citric.acid)) + 
  geom_boxplot() 
summary(redwine$citric.acid)
cor(redwine$quality,redwine$citric.acid)
```

There seems to be less correlation with citric acid and quality than with
alcohol and quality. That's because even in the higher qualitie samples like 
7 or 8, there are sizable samples that have small amounts of citric acid, 
according to the scatter plot. However, according to the box plot, the median 
of the citric acid is increasing as the quality increases, even if by a little 
amount. This shows that there is a very little correlation, also confirmed by 
the correlation coefficient which is 0.22.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(pH,volatile.acidity)) + geom_jitter(alpha = 0.2)
ggplot(data = redwine, aes(pH,fixed.acidity)) + geom_jitter(alpha = 0.2)
cor(redwine$pH,redwine$volatile.acidity)
cor(redwine$pH,redwine$fixed.acidity)



```

pH correlates more to fixed acidity than to volatile acidity. According to the
scatterplot for fixed acidity, there is a clear negative correlation - as pH
goes down, the fixed acidity level also goes down. The correlation coefficient
of fixed acidity is -0.683. I think it's interesting that fixed acidity shows
little correlation with pH and if it does, it is that higher pH increases 
volatile acidity, as shown by the medium to high amounts of volatile acidity 
in samples with pH greater than 3.4.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality, pH)) + geom_jitter(alpha = 0.2)
ggplot(data = redwine, aes(group=quality, x = quality, y =  pH)) + 
  geom_boxplot() 
summary(redwine$pH)
cor(redwine$quality,redwine$pH)


```

The points seem to be evenly dispersed vertically in each of the quality levels
for the scatterplot. This indicates a weak correlation between pH and quality.
This is confirmed by the correlation coefficient which is just -0.058.
Furthermore, the medians are all very similar to each other across quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality, total.sulfur.dioxide)) + 
  geom_jitter(alpha = 0.1)
ggplot(data = redwine, aes(group=quality, x = quality, 
                           y =  total.sulfur.dioxide)) + geom_boxplot() 
cor(redwine$quality,redwine$total.sulfur.dioxide)

```

According to the scatter plot, the "bars" made by Qualities of 5 and 6 are the 
tallest. They start decreasing in height along the ends. Perhaps we can infer
that to make an average (Quality 5 or Quality 6) wine, you can put more Sulfur
Dioxide in it. The Quality 8 wine samples have very low total sulfur dioxide 
amounts, as confirmed by the negative correlation coefficient (-0.185).



```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality, chlorides)) + geom_jitter(alpha = 0.1)
ggplot(data = redwine, aes(group=quality, x = quality, y =  chlorides)) + 
  geom_boxplot() 
cor(redwine$quality,redwine$chlorides)
```

According to the scatter plot, as Quality improves, there is very little 
difference in the general amount of chloride. The boxplot reinforces this 
because the medians are all very close to each other with the amount of 
chloride.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality, residual.sugar)) + geom_jitter(alpha = 0.1)
ggplot(data = redwine, aes(group=quality, x = quality, y =  residual.sugar)) + 
  geom_boxplot()
cor(redwine$quality,redwine$residual.sugar)
```

Similar to chloride, there seems to be little correlation of residual sugar and
quality. The points are all dispersed similarly in each quality level, showing
no cause and effect relationship between residual sugar and quality (Since 
there are more samples with 5 and 6, it is only heavier in the bottom for those
qualities.) In the box plots, the medians are all similar and so are the IQRs.
Furthermore, the correlation coefficient is very close to 0.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(quality,sulphates)) + geom_jitter(alpha = 0.2) 
ggplot(data = redwine, aes(group=quality, x = quality, y =  sulphates)) +
  geom_boxplot() 
summary(redwine$sulphates)
cor(redwine$quality,redwine$sulphates)

```

There's a slight positive correlation with sulphates and quality. As the quality
is increasing, the points are going upward. Furthermore, the boxes in the 
boxplot are also moving upward as quality increases. And a correlation 
coefficient of 0.25 is solid enough to indicate some positive relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(residual.sugar, chlorides)) + 
  geom_jitter(alpha = 0.1) + 
#Removed the top 10% of points from chlorides and residual.sugar because they 
  #seemed to be of very high values compared to the rest of the data points.
xlim(0,quantile(redwine$residual.sugar,0.9)) + 
  ylim(0,quantile(redwine$chlorides,0.9))
cor.test(redwine$residual.sugar, redwine$chlorides)

```

Lastly, let's look at the relationship between chlorides and residual sugar 
as we mentioned in the first section that their distributions look similar. In
the scatterlot, clearly, there is little correlation between the amount of 
residual sugar and chlorides since the points are all scattered. Furthermore, 
the correlation coefficient is just 0.05.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I think the relationship between alcohol and quality, sulphates and quality, 
and citric acid and quality were the most interesting. These were the top 3 
strongest predictors of quality, with alcohol being on top by a clear margin.
I think if we were to make a model, alcohol would be the strongest variable 
in that model followed by sulphates then citric acid. I was expecting pH to
have a stronger correlation with the quality of wine, but it had very little 
correlation. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The relationhip between pH and fixed acidity was very strong, which was 
expected. However, I was expecting the relationship between pH and volatile 
acidity to be stronger and negative, similar to fixed acidity and pH. 

### What was the strongest relationship you found?

The relationship of pH and fixed acidity was the strongest. However, if we're
talking about the Feature of interest, Quality, then Quality and Alcohol was 
the strongest. 

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}
#box and whisker, quality, alcohol/ density box plot
ggplot(data = redwine, aes(quality, citric.acid/sulphates)) + 
  geom_boxplot(aes(group=quality))
ggplot(data = redwine, aes(x = sulphates, y = citric.acid, color = factor(quality))) + 
  geom_point(alpha = 0.8, size = 1) + geom_smooth(method="lm",se=FALSE,size=1) + 
  scale_color_brewer(type = 'seq',guide=guide_legend(title="Quality"))

```

The scatterplot of sulphates and citric acid shows us that too much or too 
little sulphates give a bad effect to the quality. There's also a sweet spot for
the amount of citric acid that should be there. I think the ideal amount of 
sulphates and citric acid should be somewhere near the light blue patch with
sulphates being 0.75 and citric acid content being 0.50. I think if a wine 
sample has its sulphates and citric acid level in this area or even with citric
acid values as low as 0.25 (but not lower than that), that would be ideal.
According to the box plot, as the ratio of citric acid/sulphates gets higher, 
the quality increases. In other words, if the citric.acid to sulphates ratio is 
0.5, there's a good chance of the wine sample being of 7 or 8 quality.

```{r echo=FALSE}
ggplot(data = redwine, aes(quality, citric.acid/alcohol)) + 
  geom_boxplot(aes(group=quality))
ggplot(data = redwine, aes(x = alcohol, y = citric.acid, color = factor(quality))) + 
  geom_point(alpha = 0.8, size = 1) + geom_smooth(method="lm",se=FALSE,size=1) + 
  scale_color_brewer(type = 'seq',guide=guide_legend(title="Quality"))
```

This scatterplot is interesting because it shows that extremes of either alcohol
or citric acid without enough of the other will result in bad quality. This is 
because most points either too down or too to the left on the graph are darker
colors, indicating a bad quality. The top right portion of points have a 
cluster of points that are light colored, indicating a good quality. 
The boxplot confirms this by indicating that when the ratio of citric acid to 
alcohol is higher (just below 0.04) it leads to greater quality then when the 
ratio is closer to 0.

```{r echo=FALSE}
ggplot(data = redwine, aes(quality, alcohol/sulphates)) + 
  geom_boxplot(aes(group=quality))
ggplot(data = redwine, aes(x = sulphates, y = alcohol, color = factor(quality))) + 
  geom_point(alpha = 0.8, size = 1) + geom_smooth(method="lm",se=FALSE,size=1) + 
  scale_color_brewer(type = 'seq',guide=guide_legend(title="Quality"))

```


Again, this scatterplot is similar to the previous one in that when there is 
too much of either substance (alcohol or sulphates) without the other, it 
results in lower quality. Most of the lighter colored dots are in the top right
region again. It appears that the ideal sulphate level is around 0.75-1 and the 
alcohol level is around 12-13.
The boxplot makes it seem as though the ratio does not make a big difference in 
quality. Maybe because dividing the alcohol content (a big number) by the 
sulphates content (very small number) makes the ratios end up all similar to 
each other, so it's hard to look for a trend/relationship.

```{r echo=FALSE}
ggplot(data = redwine, aes(quality, pH/fixed.acidity)) + 
  geom_boxplot(aes(group=quality))
ggplot(data = redwine, aes(x = pH, y = fixed.acidity, color = factor(quality))) + 
  geom_point(alpha = 0.8, size = 1) + geom_smooth(method="lm",se=FALSE,size=1) + 
  scale_color_brewer(type = 'seq',guide=guide_legend(title="Quality"))

```

According to this scatterplot, the lighter points are more concentrated in the 
middle than the ends, where pH is about 3.3 and fixed.acidity is near 8. This
shows that a medium level of fixed acidity or pH is ideal. The box plot however 
shows that there is a slight decreasing of the ph to fixed acidity ratio as the
quality increases.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

There seemed to be a relationship between citric acids, sulphates, and quality,
as seen by the first plot. This relationship showed a sweetspot regarding the 
amount of citric acid and the amount of sulphates there should be in order to 
maximize the quality. For sulphates, too low or too much had a clear bad affect
on quality, indicating darker dots. The ideal amount seemed to be 0.75 and for
citric acid it seemed to be 0.5. I thought the patch of lighter colored blue
dots in the top right area is really where quality was a little better. I 
noticed that extremes (low or high) of most chemicals caused an adverse effect
on the quality. I thought the complentary relationship between citric acid
and alcohol was interesting because it shows that too much of either of them 
without the other can have an adverse effect but if you have a good dose of both
of them together then it leaves a good impact on quality.


### Were there any interesting or surprising interactions between features?

I thought it was interesting how although very high levels of alcohol did have
a bad effect on quality, the bad effect in general of higher levels of alcohol 
wasn't as noticeable as the bad effect of high levels of other chemicals. I
think that this may indicate that alcohol content is one of the more important
indicators of quality of red wine.


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = redwine, aes(quality)) + geom_histogram(bins = 20) +
  ggtitle("Frequency of Given Quality in Red Wine Samples") + theme(plot.title = element_text(hjust = 0.5)) + ylab("Frequency") + 
  xlab("Quality (Score between 0 and 10)")

```

### Description One

The frequency of quality is unimodal because there is a single peak of frequency
which occurs around the middle qualities (5 and 6).Both ends (3 and 8) have 
barely any scores, in other words, there's a big gap between the sum of 3 and 8 
qualities and the sum of 5 and 6 qualities. If there were more samples with 
3s and 8s, then perhaps we would be able to have a better idea of what makes 
good wine samples good and bad wine samples bad.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data = redwine, aes(quality,alcohol)) + geom_jitter(alpha = 0.2) + 
  ggtitle("Alcohol Content by Quality") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("Quality (Score between 0 and 10)") + 
  ylab("Alcohol Content (% by Volume)")

```

### Description Two

As the Quality increases, more dots are shifting upwards in the alcohol content
scale. There is a big density of dots for quality 5 volume 9% and this dense
area is dispersed more evenly in quality 6. This indicates more alcohol content
than 9% is likelier to give better quality.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = redwine, aes(quality, citric.acid/sulphates)) + 
  geom_boxplot(aes(group=quality)) + 
  ggtitle("Citric Acid / Sulphates by Quality") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("Quality (Score between 0 and 10)") + 
  ylab("Citric Acid (g / dm^3) / Sulphates (g / dm^3)")

```

### Description Three

As quality increases, the ratio of citric acid by sulphates acid is also 
increasing, moving closer to 1. This indicates that when citric acids and 
sulphate acid quantities are both closer to each other, that's when there is 
better quality. Not only do the medians get higher but the top of the boxes
(The 75th quartiles) also get bigger, so that means the trend is clear.

------

# Reflection

I think it's interesting how important of a role alcohol played in the quality
of red wine samples. More than any other factor, alcohol played the greatest 
role in determining the quality. Although it's a shame that there were not more
samples of 8 in order to see what makes them so good, I still think that 
the findings suggest that alcohol is the strongest factor for quality. Besides 
alcohol, I think citric acid also plays a strong role in the quality of red 
wine. Both the bivariable and multivariable analysis proved this fact.
I think this study could have been improved if there were more samples of 3s 
and 8s. It's unfortunate that the samples of quality were concentrated in the 
middle qualities of 5 and 6. This makes it more difficult to tell what makes 
good wine good and bad wine bad. This study should be done again with more
bimodal sampling of wines, regarding their qualities. This way we can tell what
makes good wine good and bad wine bad more clearly. I think what went successful
with this study is the fact that I was able to narrow down alcohol's 
importance as a predictor of quality of red wine. Even citric acid was narrowed
down as a factor, so that's another point of success. I think what was cause 
for struggle was that as mentioned before, there were not enough samples for 
very bad qualities and very good qualities. Another struggle was that since 
Quality is the only categorical variable, I wasn't able to make a more 
variety of charts for the analysis. Instead, I was forced to always use 
Quality as the categorical variable in the analysis.